---
subtitle: "3-GSEA analysis comparing TGFbRII-KO to control with bulkRNAseq data"
bibliography: bib.bib
link-citations: yes
date: "Rendered `r format (Sys.time(), format='%Y-%m-%d %H:%M:%S %z')`"
output: 
  pdf_document:
    pandoc_args: 
      - '../../common.yaml'
      - --listings
    includes: 
      in_header: '../../preamble.tex'
    toc: true
    number_sections: true
    fig_width: 7
    fig_height: 6
    fig_caption: true
    df_print: tibble
---


```{r include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80),
               tidy=TRUE,
               message = FALSE, 
               warning = FALSE)

```

\newpage


# Description

The fold change between TGFbRII-KO and control for each gene (logFC) calculated in previous step was used to rank the genes from the most up-regulated to the most down-regulated gene in TGFbRII-KO. The ranked genelist was then subjectted to Gene Set Enrichment Analysis (GSEA). The genesets in MSigDB used for GSEA were extracted from msigdbr package [@msigdbr] and then used by `GSEA` function from `clusterProfiler` pakcage [@clusterProfiler]. `ggplot2` package was used to build plots [@ggplot2]. 


# Prepare data for GSEA analysis

Load packages:

```{r}
suppressMessages(library(ggplot2))
suppressMessages(library(clusterProfiler))
suppressMessages(library(msigdbr))
```

Load data: 

```{r eval=FALSE}
dds <- readRDS("../1_Bulk_RNAseq_Data_preparation/dds.Rds")

# We will use the logFC calculated in previous step for ranking the genes: 
DE.res.KO_vs_WT <- read.table(
  file = "../2_Differential_Expression_TGFbRII-KO_vs_WT/DE.res.KO_vs_WT.txt", 
  sep = "\t")

genelist.ranked.KO_CTL <- DE.res.KO_vs_WT$log2FoldChange.shrunk
names(genelist.ranked.KO_CTL) <- rownames(DE.res.KO_vs_WT)
genelist.ranked.KO_CTL <- sort(genelist.ranked.KO_CTL, decreasing = TRUE)
```

Extract genesets from MSigDB database via `msigdbr` package: 

```{r eval=FALSE}
gs_c5_bp <- msigdbr(species = "Mus musculus", category = "C5", 
                    subcategory = "GO:BP") %>% dplyr::select(gs_name, gene_symbol)
gs_c5_cc <- msigdbr(species = "Mus musculus", category = "C5", 
                    subcategory = "GO:CC") %>% dplyr::select(gs_name, gene_symbol)
gs_c5_mf <- msigdbr(species = "Mus musculus", category = "C5", 
                    subcategory = "GO:MF") %>% dplyr::select(gs_name, gene_symbol)
gs_h <- msigdbr(species = "Mus musculus", category = "H") %>% 
  dplyr::select(gs_name, gene_symbol)
```

## Enrichment with C1-Hallmark genesets

```{r eval=FALSE}

gsea_h <- GSEA(geneList = genelist.ranked.KO_CTL, 
                   TERM2GENE = gs_h, 
                   verbose = FALSE, seed = TRUE)
gsea_h@result <- gsea_h[order(gsea_h$NES, decreasing = TRUE),]
write.table(x = gsea_h@result, file = "./gsea_h.txt", sep = "\t")
```


```{r include=FALSE}
gsea_h <- readRDS("./gsea_h.rds")
```

## Enrichment with C5-GO (BP) genesets

```{r eval=FALSE}
gsea_c5_bp <- GSEA(geneList = genelist.ranked.KO_CTL, 
                   TERM2GENE = gs_c5_bp, 
                   verbose = FALSE, seed = TRUE)
gsea_c5_bp@result <- gsea_c5_bp[order(gsea_c5_bp$NES, decreasing = TRUE),]
```


```{r include=FALSE}
gsea_c5_bp <- readRDS("./gsea_c5_bp.rds")
```


## Enrichment with C5-GO (MF)

```{r eval=FALSE}
gsea_c5_mf <- GSEA(geneList = genelist.ranked.KO_CTL, 
                   TERM2GENE = gs_c5_mf, 
                   verbose = FALSE, seed = 123)
gsea_c5_mf@result <- gsea_c5_mf[order(gsea_c5_mf$NES, decreasing = TRUE),]
```


```{r include=FALSE}
gsea_c5_mf <- readRDS("./gsea_c5_mf.rds")
```


## Enrichment with C5-GO (CC)

```{r eval=FALSE}
gsea_c5_cc <- GSEA(geneList = genelist.ranked.KO_CTL, 
                   TERM2GENE = gs_c5_cc, 
                   verbose = FALSE, seed = TRUE)
gsea_c5_cc@result <- gsea_c5_cc[order(gsea_c5_cc$NES, decreasing = TRUE),]
```


```{r include=FALSE}
gsea_c5_cc <- readRDS("./gsea_c5_cc.rds")
```

## C2-Pathway

```{r eval=FALSE}
gs_c2 <- msigdbr(species = "Mus musculus", 
                 category = "C2") %>% 
  dplyr::select(gs_name, gene_symbol)

gsea_c2 <- GSEA(geneList = genelist.ranked.KO_CTL, 
                   TERM2GENE = gs_c2, 
                   verbose = FALSE, seed = TRUE)
gsea_c2@result <- gsea_c2[order(gsea_c2$NES, decreasing = TRUE),]
```


```{r include=FALSE}
gsea_c2 <- readRDS("./gsea_c2.rds")
```

# Make summarising dotplot

Build dataframe with results: 

```{r}
df.pos <- do.call("rbind", list(gsea_c2[1,], # PLASARI_TGFB1_TARGETS_10HR_DN
                      gsea_c5_bp[1,],  # GOBP_CELL_CELL_ADHESION_VIA_PLASMA_MEMBRANE_ADHESION_MOLECULES
                      gsea_c5_bp["GOBP_MITOTIC_NUCLEAR_DIVISION",],
                      gsea_h["HALLMARK_INTERFERON_ALPHA_RESPONSE",], 
                      gsea_h["HALLMARK_G2M_CHECKPOINT",],
                      gsea_h["HALLMARK_E2F_TARGETS",]) 
        )


df.neg <- do.call("rbind", list(gsea_c2["KARLSSON_TGFB1_TARGETS_UP",], 
                      gsea_c2["KEGG_ANTIGEN_PROCESSING_AND_PRESENTATION",], 
                      gsea_c5_bp["GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_EXOGENOUS_PEPTIDE_ANTIGEN",], 
                      gsea_c5_bp["GOBP_TISSUE_REMODELING",],
                      gsea_h["HALLMARK_MTORC1_SIGNALING",],
                      gsea_c2["FOSTER_TOLERANT_MACROPHAGE_UP",]) 
        )

# Order the data frame by NES
df.both <- rbind(df.pos, df.neg)
df.both <- df.both %>% arrange(NES)

# Ensure the Pathway factor levels follow the order in the data frame
df.both$ID <- factor(df.both$ID, levels = df.both$ID)
```


```{r fig.width=10, fig.height=5}
# Create the plot
ggplot(df.both, aes(x = NES, y = ID)) +
  geom_point(aes(color = qvalue, size = -log10(p.adjust)), alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_gradient(low = "red", high = "blue") +
  labs(x = "NES", y = "", title = "GSEA Results Summary") +
  theme_minimal() +
  theme(axis.text.y = element_text(hjust = 1)) + 
  guides(size = guide_legend(title = "-log(p.adjust)"), color = guide_colorbar(title = "FDR q.val"))
```


# Session information

R sesssion: 

```{r}
sessionInfo()
```

# References

